Reactive SINDy: Discovering governing reactions from concentration data

### 2018-12-08 docs/manuscript_jcp.tex
wip

We additionally demonstrate the algorithm
on time series data of the mitogen activated protein kinases (MAPK) pathway as 
an example for a bimodal system and on time series data of the
Lotka-Volterra system which describes oscillatory predator-prey dynamics subject to
social friction. In both systems reactive SINDy recovers the generating 
reaction network whereas non-sparse estimation detects many false processes.

@@

\subsection{Application to Lotka--Volterra system\label{sec:lv}}
As biological pathways often exhibit oscillatory behavior~\cite{Iqbal2008tnf} which can stem 
from positive or negative feedback loops~\cite{Shin2009positive} we apply reactive SINDy to
an idealized oscillatory system, namely the Lotka--Volterra system. 
The predator-prey dynamics of two species $\mathrm{A}$ (prey) and $\mathrm{B}$ 
(predator) is defined by the reaction network
\begin{align*}
\mathrm{A}              & \rightharpoonup \mathrm{A} + \mathrm{A} &  & \text{(prey growth)},\\
\mathrm{A} + \mathrm{B} & \rightharpoonup \mathrm{B} + \mathrm{B} &  & \text{(predator eats prey)},\\
\mathrm{B}              & \rightharpoonup \emptyset &  & \text{(predator decay)},\\
\mathrm{A} + \mathrm{A} & \rightharpoonup \emptyset &  & \text{(prey friction)},\\
\mathrm{B} + \mathrm{B} & \rightharpoonup \emptyset &  & \text{(predator friction)}.
\end{align*}
From this model we generate concentration time series data with $200$ frames 
which is displayed in Fig.~\ref{fig:lv}a. We construct the library $\Theta$ 
of candidate reactions given in Tab.~\ref{tab:reaction-library-lv}. 
We determine the hyperparameter $\alpha=2.7\times10^{-7}$ by shuffling 
the data and performing 5-fold cross validation. We obtain the estimated 
rate constants by solving the minimization problem (\ref{method:minimizationproblem}) 
with $\lambda=1$. The results are given in Fig.~\ref{fig:lv}b.
Least-squares estimation detects all reactions of the ground truth model but also
two false processes ($\theta_{6}$ and $\theta_{7}$) with a higher rate than
the two first true processes ($\theta_{1}$ and $\theta_{2}$). Reactive SINDy detects
the true reaction network with minor deviations in rate constants. As in 
Sec.~\ref{sec:mapk} we consider the $L_1$ norm of the relative distance to the ground truth
for non-zero rate constants
\begin{equation*}
\sum_{r=1}^{5} \left|(\hat{\xi}_{r} - \xi_r)/\xi_r\right|
\end{equation*}
which yields $75\%$ error for least-squares and $7\%$ error for reactive SINDy.

\begin{figure}
    \centering
    \includegraphics[width=\singlecolumnwidth]{figures_jcp/7.pdf}
    \caption{\label{fig:lv}Application of reactive SINDy to the Lotka-Volterra system with social friction. \textbf{(a)}~Concentration data as a function of time for predator and prey species. \textbf{(b)}~Estimated rate coefficients of candidate reactions (see Tab.~\ref{tab:reaction-library-lv}) after application of reactive SINDy (regularized) to the time series data. Least-squares estimation (LSQ) estimation and the ground truth model for comparison.}
\end{figure}

@@

In this work we have extended the SINDy method to reactive SINDy,
not only parsimoniously detecting potentially nonlinear terms in a
dynamical system from noisy data, but also yielding, in this case,
a sparse set of rates with respect to generating reactions (\ref{method:the-reactions}).
Mathematically this has been achieved by permitting vector-valued
basis functions and obtaining a tensor linear regression problem.
We have applied this method on data generated from a gene regulation
network, a MAPK pathway, and a Lotka-Volterra system and 
could successfully recover the underlying reaction networks.

@@

\begin{table*}[h!]
    \centering %
    %\resizebox{.9\textwidth}{!}{%
    \begin{tabular}{rclcc}
        \toprule 
        \multicolumn{3}{c}{Reaction} & rate & description \tabularnewline
        \midrule 
        $\mathrm{A} + \mathrm{A}$ & $\rightharpoonup$ & $\emptyset             $  & $k_{1} =0.1$ & social friction of prey\tabularnewline
        $\mathrm{B} + \mathrm{B}$ & $\rightharpoonup$ & $\emptyset             $  & $k_{2} =0.1$ & social friction of predator\tabularnewline
        $\mathrm{A}             $ & $\rightharpoonup$ & $\mathrm{A} + \mathrm{A}$ & $k_{3} =1  $ & prey growth\tabularnewline
        $\mathrm{A} + \mathrm{B}$ & $\rightharpoonup$ & $\mathrm{B} + \mathrm{B}$ & $k_{4} =1  $ & predator eats prey\tabularnewline
        $\mathrm{B}             $ & $\rightharpoonup$ & $\emptyset             $  & $k_{5} =1  $ & predator decays\tabularnewline
        $\mathrm{A} + \mathrm{B}$ & $\rightharpoonup$ & $\mathrm{A} + \mathrm{A}$ & $k_{6} =0  $ & artificial reaction\tabularnewline
        $\mathrm{A}             $ & $\rightharpoonup$ & $\emptyset             $  & $k_{7} =0  $ & artificial reaction\tabularnewline
        $\mathrm{B} + \mathrm{B}$ & $\rightharpoonup$ & $\mathrm{B}             $ & $k_{8} =0  $ & artificial reaction\tabularnewline
        $\mathrm{B}             $ & $\rightharpoonup$ & $\mathrm{B} + \mathrm{B}$ & $k_{9} =0  $ & artificial reaction\tabularnewline
        $\mathrm{A} + \mathrm{A}$ & $\rightharpoonup$ & $\mathrm{A}             $ & $k_{10}=0  $ & artificial reaction\tabularnewline
        $\mathrm{A} + \mathrm{B}$ & $\rightharpoonup$ & $\mathrm{A}             $ & $k_{11}=0  $ & artificial reaction\tabularnewline
        $\mathrm{A} + \mathrm{B}$ & $\rightharpoonup$ & $\mathrm{B}             $ & $k_{12}=0  $ & artificial reaction\tabularnewline
        $\mathrm{A} + \mathrm{A}$ & $\rightharpoonup$ & $\mathrm{B}             $ & $k_{13}=0  $ & artificial reaction\tabularnewline
        $\mathrm{A}             $ & $\rightharpoonup$ & $\mathrm{B}             $ & $k_{14}=0  $ & artificial reaction\tabularnewline
        $\mathrm{B}             $ & $\rightharpoonup$ & $\mathrm{A}             $ & $k_{15}=0  $ & artificial reaction\tabularnewline
        $\mathrm{A}             $ & $\rightharpoonup$ & $\mathrm{B} + \mathrm{B}$ & $k_{16}=0  $ & artificial reaction\tabularnewline
        \bottomrule
    \end{tabular}\caption{\label{tab:reaction-library-lv}Full set of ansatz reactions $\Theta$ used in Sec.~\ref{sec:lv} for the Lotka-Volterra system. The given rate constants define the ground truth reaction model.}
\end{table*}

>>>>

We additionally demonstrate the algorithm
on time series data of the mitogen activated protein kinases (MAPK) pathway as 
an example for a bimodal system and on time series data of the
Lotka--Volterra system which describes oscillatory predator-prey dynamics subject to
social friction. In both systems reactive SINDy recovers the generating 
reaction network whereas non-sparse estimation detects many false processes.

@@

\subsection{Application to Lotka--Volterra system\label{sec:lv}}
As biological pathways often exhibit oscillatory behavior~\cite{Iqbal2008tnf} which can stem 
from positive or negative feedback loops~\cite{Shin2009positive} we apply reactive SINDy to
an idealized oscillatory system, namely the Lotka--Volterra system. 
The predator-prey dynamics of two species $\mathrm{A}$ (prey) and $\mathrm{B}$ 
(predator) is defined by the reaction network
\begin{align*}
\mathrm{A}              & \rightharpoonup \mathrm{A} + \mathrm{A} &  & \text{(prey growth)},\\
\mathrm{A} + \mathrm{B} & \rightharpoonup \mathrm{B} + \mathrm{B} &  & \text{(predator eats prey)},\\
\mathrm{B}              & \rightharpoonup \emptyset &  & \text{(predator decay)},\\
\mathrm{A} + \mathrm{A} & \rightharpoonup \emptyset &  & \text{(prey friction)},\\
\mathrm{B} + \mathrm{B} & \rightharpoonup \emptyset &  & \text{(predator friction)}.
\end{align*}
From this model we generated concentration time series data with $200$ frames 
which is displayed in Fig.~\ref{fig:lv}a. The library 
of ansatz reactions $\Theta$ is given in Tab.~\ref{tab:reaction-library-lv}. 
The hyperparameter $\alpha=2.7\times10^{-7}$ was determined by shuffling 
the data and performing 5-fold cross validation. The estimated 
rate constants were obtained by solving the minimization problem (\ref{method:minimizationproblem}) 
with $\lambda=1$. The results are depicted in Fig.~\ref{fig:lv}b.
Least-squares estimation detects all reactions of the ground truth model but also
two spurious processes ($\theta_{6}$ and $\theta_{7}$) with a higher rate than
the first two underlying processes ($\theta_{1}$ and $\theta_{2}$). Reactive SINDy recovers
the true reaction network with minor deviations in rate constants. As in 
Sec.~\ref{sec:mapk}, considering the $L_1$ norm of the relative distance to the ground truth
for non-zero rate constants
\begin{equation*}
\sum_{r=1}^{5} \left|(\hat{\xi}_{r} - \xi_r)/\xi_r\right|
\end{equation*}
yields $75\%$ error for least-squares and $7\%$ error for reactive SINDy.

\begin{figure}
    \centering
    \includegraphics[width=\singlecolumnwidth]{figures_jcp/7.pdf}
    \caption{\label{fig:lv}Application of reactive SINDy to the Lotka--Volterra system with social friction. \textbf{(a)}~Concentration data as a function of time for predator and prey species. \textbf{(b)}~Estimated rate coefficients of candidate reactions (see Tab.~\ref{tab:reaction-library-lv}) after application of reactive SINDy (regularized) to the time series data. Least-squares estimation (LSQ) estimation and the ground truth model for comparison.}
\end{figure}

@@

In this work we have extended the SINDy method to reactive SINDy,
not only parsimoniously detecting potentially nonlinear terms in a
dynamical system from noisy data, but also yielding, in this case,
a sparse set of rates with respect to generating reactions (\ref{method:the-reactions}).
Mathematically this has been achieved by permitting vector-valued
basis functions and obtaining a tensor linear regression problem.
We have applied this method on data generated from a gene regulation
network, a MAPK pathway, and a Lotka--Volterra system and 
could successfully recover the underlying reaction networks.

@@

\begin{table*}[h!]
    \centering %
    %\resizebox{.9\textwidth}{!}{%
    \begin{tabular}{rclcc}
        \toprule 
        \multicolumn{3}{c}{Reaction} & rate & description \tabularnewline
        \midrule 
        $\mathrm{A} + \mathrm{A}$ & $\rightharpoonup$ & $\emptyset             $  & $k_{1} =0.1$ & social friction of prey\tabularnewline
        $\mathrm{B} + \mathrm{B}$ & $\rightharpoonup$ & $\emptyset             $  & $k_{2} =0.1$ & social friction of predator\tabularnewline
        $\mathrm{A}             $ & $\rightharpoonup$ & $\mathrm{A} + \mathrm{A}$ & $k_{3} =1  $ & prey growth\tabularnewline
        $\mathrm{A} + \mathrm{B}$ & $\rightharpoonup$ & $\mathrm{B} + \mathrm{B}$ & $k_{4} =1  $ & predator eats prey\tabularnewline
        $\mathrm{B}             $ & $\rightharpoonup$ & $\emptyset             $  & $k_{5} =1  $ & predator decays\tabularnewline
        $\mathrm{A} + \mathrm{B}$ & $\rightharpoonup$ & $\mathrm{A} + \mathrm{A}$ & $k_{6} =0  $ & artificial reaction\tabularnewline
        $\mathrm{A}             $ & $\rightharpoonup$ & $\emptyset             $  & $k_{7} =0  $ & artificial reaction\tabularnewline
        $\mathrm{B} + \mathrm{B}$ & $\rightharpoonup$ & $\mathrm{B}             $ & $k_{8} =0  $ & artificial reaction\tabularnewline
        $\mathrm{B}             $ & $\rightharpoonup$ & $\mathrm{B} + \mathrm{B}$ & $k_{9} =0  $ & artificial reaction\tabularnewline
        $\mathrm{A} + \mathrm{A}$ & $\rightharpoonup$ & $\mathrm{A}             $ & $k_{10}=0  $ & artificial reaction\tabularnewline
        $\mathrm{A} + \mathrm{B}$ & $\rightharpoonup$ & $\mathrm{A}             $ & $k_{11}=0  $ & artificial reaction\tabularnewline
        $\mathrm{A} + \mathrm{B}$ & $\rightharpoonup$ & $\mathrm{B}             $ & $k_{12}=0  $ & artificial reaction\tabularnewline
        $\mathrm{A} + \mathrm{A}$ & $\rightharpoonup$ & $\mathrm{B}             $ & $k_{13}=0  $ & artificial reaction\tabularnewline
        $\mathrm{A}             $ & $\rightharpoonup$ & $\mathrm{B}             $ & $k_{14}=0  $ & artificial reaction\tabularnewline
        $\mathrm{B}             $ & $\rightharpoonup$ & $\mathrm{A}             $ & $k_{15}=0  $ & artificial reaction\tabularnewline
        $\mathrm{A}             $ & $\rightharpoonup$ & $\mathrm{B} + \mathrm{B}$ & $k_{16}=0  $ & artificial reaction\tabularnewline
        \bottomrule
    \end{tabular}\caption{\label{tab:reaction-library-lv}Full set of ansatz reactions $\Theta$ used in Sec.~\ref{sec:lv} for the Lotka--Volterra system. The given rate constants define the ground truth reaction model.}
\end{table*}
